# Adopted Pet Analysis - Research Question 1

### EDA Part 1 - Z-Tests for Proportion\

#### Read in the file for us_all_jantodec_adopted_dogsANDcats file
```{r}
library(dplyr)
data <- read.csv(file = "/Users/debbiesubocz/uw-code/BIOST557/Project-R/data/us_all_jantodec_adopted_dogsANDcats.csv")
data <- data %>%
  filter(contact.address.country == "US")
head(data, 5)
```
#### Determine the states include in the dataset

```{r}
print(unique(data$contact.address.state))
unique_states_count <- data %>%
  filter(type == "Dog") %>%
  distinct(contact.address.state) %>%
  nrow()

print(unique_states_count)
```

#### Determine the time range of the dataset

```{r}
oldest_published <- min(data$published_at)
most_recent_published <- max(data$published_at)

oldest_status_changed <- min(data$status_changed_at)
most_recent_status_changed <- max(data$status_changed_at)


print(paste0("published_at date range: ", oldest_published, " to ", most_recent_published))
print(paste0("status_changed_at date range: ", oldest_status_changed, " to ", most_recent_status_changed))

```

#### Determine the statuses in the dataset
```{r}
status <- data %>% 
  group_by(status, type) %>%
  summarise(Adopted_Pets = n())

head(status, 5)
```
#### Determine the breeds in the dataset

```{r}
unique_dog_breeds_count <- data %>%
  filter(type == "Dog") %>%
  distinct(breed) %>%
  nrow()

dog_breeds <- data %>% 
  group_by(breed) %>%
  filter(type == 'Dog') %>%
  summarise(Adopted_Pets = n())

print(head(dog_breeds, 5))
print(paste0("Unique dog breeds count: ",unique_dog_breeds_count))


```


#### Group by Animal Type
```{r}
sample_size <- data %>% 
  group_by(type) %>%
  summarise(Adopted_Pets = n())

head(sample_size, 5)
```
#### Group by Animal Type, By State
```{r}
eda_data <- data %>% 
  group_by(contact.address.state, type) %>%
  summarise(Adopted_Pets = n())

head(eda_data, 5)
```

```{r}
boxplot(Adopted_Pets ~ type, data = eda_data, main = "Range of Counts of Adopted Dogs and Cats per US State", xlab = "Type", ylab = "Count of Adopted Pets")
```

#### Get the proportion of dogs and cats for each state
```{r}
# Calculate the total count for each state and type
totals <- eda_data %>%
  group_by(contact.address.state) %>%
  summarise(total_count = sum(Adopted_Pets))

head(totals, 5)

# Join the totals back to the original data frame and calculate the proportion
by_state <- eda_data %>%
  left_join(totals, by = c("contact.address.state"))

by_state <- by_state %>%
  mutate(Prop = Adopted_Pets / total_count)

head(by_state, 5)

```

#### Export the dataframe to a csv (for upload to Tableau for heatmap creation)
```{r}
write.csv(by_state, "props_for_heatmap_updated.csv")
```

#### Create bins for PNW, SC, and NE regions

```{r}
by_region <- by_state

by_region$region <- ifelse(by_region$contact.address.state %in% c("WA", "OR", "ID"), "PNW", 
                 ifelse(by_region$contact.address.state %in% c("CT", "ME", "MA", "NH", "NJ", "NY", "PA", "RI", "VT"), "NE", 
                        ifelse(by_region$contact.address.state %in% c("KY", "TN", "AL", "MS","AR", "LA", "OK", "TX"), "SC", NA)))

by_region <- by_region %>%
  filter(region %in% c('PNW', 'NE', 'SC'))

head(by_region, 5)
```

```{r}
boxplot(Prop ~ region + type, data = by_region, main = "Range of Proportions of Adopted Dogs and Cats per US State by Region", xlab = "Region, Type", ylab = "Proportion of Adopted Pets")
```



### Statistical Tests - Part 1

#### One Sample Z-Test for a Proportion

```{r}
by_type <- data %>% 
  group_by(type) %>%
  summarise(Count = n())

by_type <- by_type %>%
  mutate(Prop = Count / sum(Count))

print(by_type)
```

```{r}
test_result <- prop.test(x = by_type$Count[2], n = sum(by_type$Count), p = 0.5, correct = FALSE)
print(test_result)

cat("z test statistic:", sqrt(test_result$statistic), "\n")
cat("p-value:", test_result$p.value, "\n")
```

#### Two Sample Z-Test for Comparing Two Proportions

```{r}
by_region <- data %>% 
  group_by(contact.address.state, type) %>%
  summarise(Count = n())

by_region$region <- ifelse(by_region$contact.address.state %in% c("WA", "OR", "ID"), "PNW", 
                 ifelse(by_region$contact.address.state %in% c("CT", "ME", "MA", "NH", "NJ", "NY", "PA", "RI", "VT"), "NE", 
                        ifelse(by_region$contact.address.state %in% c("KY", "TN", "AL", "MS","AR", "LA", "OK", "TX"), "SC", NA)))

by_region <- by_region %>% 
  group_by(type, region) %>%
  filter(region %in% c('PNW', 'NE', 'SC')) %>%
  summarize(Adopted_Count = sum(Count))

head(by_region, 5)
```

```{r}
# Create a table of counts for dogs and cats in each region
table <- xtabs(Adopted_Count ~ type + region, data = by_region)

# Calculate the proportions of dogs in each region
prop_dog_PNW <- table["Dog", "PNW"] / sum(table[, "PNW"])
prop_dog_SC <- table["Dog", "SC"] / sum(table[, "SC" ])

print(table)
cat("\nProportion of dogs in PNW: ", prop_dog_PNW, "\n")
cat("Proportion of dogs in SC: ", prop_dog_SC)
```



```{r}
# Perform a two-sample z-test for comparing two proportions
results <- prop.test(x = c(table["Dog", "PNW"], table["Dog", "SC"]), 
          n = c(sum(table[, "PNW"]), sum(table[, "SC"])),
          correct = FALSE)
print(results)
cat("z test statistic:", sqrt(results$statistic), "\n")
cat("p-value:", results$p.value, "\n")
```
### EDA Part 2 - Welch Two-Sample Test with t-distribution\

#### Read in the file for adopted_pets_per_month file

```{r}
data_2 <- read.csv(file = "/Users/debbiesubocz/uw-code/BIOST557/Project-R/data/adopted_pets_per_month.csv")
data_2 <- data_2 %>%
  filter(!(month_year %in% c("2023-03", "2023-02", "2023-01")))


print(head(data_2, 10))
```
#### Determine the states included in the dataset

```{r}
print(unique(data_2$contact.address.state))
unique_states_count <- data_2 %>%
  distinct(contact.address.state) %>%
  nrow()

print(unique_states_count)
```

#### Determine the time range of the dataset

```{r}
oldest <- min(data_2$month_year)
most_recent <- max(data_2$month_year)

unique_months_count <- data_2 %>%
  distinct(month_year) %>%
  nrow()

print(paste0("published_at date range: ", oldest, " to ", most_recent))
print(paste0("number of unique months: ", unique_months_count))

```

#### Determine if 24 months is a large enough sample size
```{r}
tx <- data_2 %>%
  filter(contact.address.state == "TX")
wa <- data_2 %>%
  filter(contact.address.state == "WA")
sd_tx = sd(tx$Count.of.Adopted.Cats)
sd_wa = sd(wa$Count.of.Adopted.Dogs)

sd_pooled = sqrt(((sd_tx)^2 + (sd_wa) ^2) /2)
print(paste0("sd_pooled: ",sd_pooled))

diff_in_means = 96
d = diff_in_means / sd_pooled
print(paste0("difference in means to calc. effect size: ", diff_in_means))
print(paste0("d for use in pwr.t.test: ", d))
```

```{r}
library(pwr)
pwr.t.test(n = NULL, 
           d = d , 
           sig.level = 0.05, 
           power = 0.75, 
           type = "two.sample", 
           alternative = "two.sided")
```

#### Create boxplot

```{r}
library(tidyr)
data_2_renamed <- data_2 %>%
  rename(Cat = Count.of.Adopted.Cats, Dog = Count.of.Adopted.Dogs)

df_long <- pivot_longer(data_2_renamed, cols = c("Cat", "Dog"), 
                        names_to = "Type", values_to = "Count_of_Adopted_Pets")
print(head(df_long, 10))
```

```{r}
boxplot(Count_of_Adopted_Pets ~ Type + contact.address.state, data = df_long, main = "Range of Counts of Adopted Pets per Month by State", xlab = "Type, State", ylab = "Count of Adopted Pets")
```


### Statistical Tests - Part 2

#### Welch Two Sample t-test


```{r}
tx <- data_2 %>% 
  filter(contact.address.state %in% c('TX'))

wa <- data_2 %>% 
  filter(contact.address.state %in% c('WA'))

print(tx)
print(wa)
```



```{r}
# Conduct the two-sample t-test
results_t_test <- t.test(
  x = tx$Count.of.Adopted.Dogs, 
  y = wa$Count.of.Adopted.Dogs,
  var.equal = FALSE
)

print(results_t_test)
```

